---
title: "Calculating conditional of phat"
author: "Siyuan Ma"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
  df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../")
```

```{r setup2, include=FALSE}
library(magrittr)
library(ggplot2)
smar::sourceDir("R/")
set.seed(1)
```

```{r load data, include=FALSE}
load("../data/genera.RData")
physeq <- physeq_genera %>% 
  phyloseq::subset_samples(dataset_name == "MucosalIBD") %>% 
  smar::prune_taxaSamples(flist_taxa = genefilter::kOverA(k = 5, A = 1))
mat_X_count <- smar::otu_table2(physeq)
mat_X_p <- apply(mat_X_count, 2, function(x) x / sum(x))
```

* We start with a model on $N$, $f_N(N_1, \dots, N_m)$ with copula:
  \begin{align}
    f_N(N_1, \dots, N_m) 
      &= \left(\prod f_N(N_j)\right) \times f_u(u_1, \dots, u_m) \\
      &= \left(\prod f_N(N_j)\right) \times \frac{f_g(F^{-1}_g(u_1), \dots, F^{-1}_g(u_m))}
         {\prod f_g(F^{-1}_g(u_j))}
  \end{align}
  * $u_j = F_N(N_j)$ is the percentile of $N_j$
  * $f_g(g_1, \dots, g_m)$ is multivariate Gaussian

* For generating M-H samples,
  \begin{align}
    f_{\hat{p}}(\hat{p}_j | \hat{p}_{-j}/\sum \hat{p}_{-j})
      &\propto f_{\hat{p}}(\hat{p}_1, \dots, \hat{p}_m) \\
      &= \int f_N(R\hat{p}_1, \dots, R\hat{p}_m) R^{m - 1} dR
  \end{align}
  * $R = \sum N$ (total read depth)

* Plugging in $f_N$
  \begin{align}
    f_N(R\hat{p}_1, \dots, R\hat{p}_m) 
      &= \left(\prod f_N(R\hat{p}_j) \right) \times 
         \frac{f_g(F_g^{-1}(F_N(R\hat{p}_1)), \dots, F_g^{-1}(F_N(R\hat{p}_m)))}
         {\prod f_g(F_g^{-1}(F_N(R\hat{p}_1)))} 
  \end{align}
  
* Observations
  * Integrand seems to suggest $F_N$ should follow some log exponential family
    distribution
    * Let $r = log(R)$, $n_j = log(N_j)$:
      * $R^{m - 1} dR = \exp(m r) d r$
      * $f_N(R\hat{p}_j) = f_n(r + log(p_j))$
  * $R\hat{p}_j$ fixed at $0$ for $\hat{p}_j = 0$
  
* Simple case: assume that $N_1, \dots, N_m = \exp\left(\text{MV Gaussian}\right)
  \times \prod I_j$, then integration is straightforward
  * Is not really reasonable though, as assumes dependency only exists for 
    non-zero counts