---
title: "One parameter setting for proof of concept CRT"
author: "Siyuan Ma"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "../", echo=FALSE)
library(magrittr)
library(ggplot2)
```

```{r steup2, include=FALSE}
dir_output <- "../results/CRT_proof/"
dir.create(dir_output, recursive = TRUE)
source("R/sample_CR.R")
future::plan(future::multisession)
```

* Setup
  
  * $X$: $p = 20, n = 100$. Params taken as top 20 abundant features in a stool fit. 5 random duplicates.
  
  * $Y\sim \text{Bernoulli}$. $\text{logit E}Y =\frac{X_6}{X_6 + X_{14}} + 2 * \frac{X_{14}}{X_6 + X_{14}}$, $\Rightarrow$ $X_6$, $X_{14}$ are TPs, rest are null. (They are the top and 11th most abundant features).
    
  * CRT: $m = 100$ random samples, step size $= 10$ during MCMC sampling. "Serial" sampling scheme.
  
  * Model: CV glm lasso. Variables are the feature being tested, and renormalized version of the rest.
  
* Results:
```{r simulate datasets, include=FALSE}
# R <- 5
# load("../data/params_export.RData")
# set.seed(1)
# tb_sim <- tibble::tibble(
#   params_a = list(params),
#   params_x = list(params %>% 
#                     SparseDOSSA2:::params_a_to_x()),
#   n = 100,
#   R = 1:R,
# ) %>% 
#   dplyr::mutate(i_sim = seq_len(dplyr::n())) %>%
#   dplyr::group_by(i_sim) %>%
#   dplyr::mutate(a_samples =
#                   SparseDOSSA2:::simulate_a(n = n, params = params_a[[1]]) %>%
#                   list(),
#                 x_samples =
#                   t(apply(a_samples[[1]], 1, function(x) x / sum(x))) %>%
#                   list()
#   ) %>%
#   dplyr::ungroup()
```

```{r simulate y, include = FALSE}
# tb_sim <- 
#   tb_sim %>% 
#   dplyr::group_by(R) %>% 
#   dplyr::mutate(covariates = x_samples[[1]][, features_TP] %>% 
#                   apply(1, MMUPHin:::TSS) %>% 
#                   t() %>% 
#                   apply(2, function(x) (x - mean(x)) / sd(x)) %>% 
#                   list(),
#                 prob = (covariates[[1]] %*% c(1, 2)) %>% 
#                   as.vector() %>% 
#                   SparseDOSSA2:::expit() %>% 
#                   list()) %>% 
#   dplyr::mutate(y = rbinom(
#     n = nrow(x_samples[[1]]),
#     size = 1,
#     prob = prob[[1]]
#   ) %>% 
#     list())
# save(tb_sim, file = paste0(dir_output, "tb_sim.RData"))
load(paste0(dir_output, "tb_sim.RData"))
features_TP <- tibble::tibble(
  mean = (1 - tb_sim$params_x[[1]]$pi0) * exp(tb_sim$params_x[[1]]$mu)
) %>%
  dplyr::mutate(feature = seq_len(dplyr::n())) %>%
  dplyr::arrange(-mean) %>%
  {.[c(1, 11), ]$feature}
features_FP <- tibble::tibble(
  mean = (1 - tb_sim$params_x[[1]]$pi0) * exp(tb_sim$params_x[[1]]$mu)
) %>%
  dplyr::mutate(feature = seq_len(dplyr::n())) %>%
  dplyr::arrange(-mean) %>%
  {.[c(2, 12), ]$feature}
```

```{r CRT sampling, echo=FALSE}
# for(r in tb_sim$R) {
#   for(i_feature in c(features_TP, features_FP))
#   {
#     samples_for_testing <- 
#       sample_CR(data = tb_sim$x_samples[[r]],
#                 ind_feature = seq_len(20) == i_feature,
#                 params_a = tb_sim$params_a[[r]],
#                 params_x = tb_sim$params_x[[r]],
#                 m = 100, space_size = 10,
#                 debug_dir = paste0(dir_output, "samples_CR_",
#                                    r, "_", i_feature, ".RData"))
#     save(samples_for_testing,
#          file = paste0(dir_output, "samples_for_testing_",
#                        r, "_", i_feature, ".RData"))
#   }
# }
```

```{r sanity check, echo=FALSE}
# mat_otu <- rbind(tb_sim$x_samples[[r]],
#                  samples_for_testing[1:5] %>% 
#                    purrr::reduce(rbind)) %>% 
#   t()
# metadata <- data.frame(filler = 1,
#                        dataset = rep(0:5,
#                                        each = nrow(tb_sim$x_samples[[r]])))
# physeq <- phyloseq::phyloseq(
#   otu_table = phyloseq::otu_table(mat_otu, taxa_are_rows = TRUE),
#   sample_data = metadata
# )
# ordination <- phyloseq::ordinate(physeq, method = "MDS")
# phyloseq::plot_ordination(physeq, ordination, color = "dataset")
```


```{r CRT fitting, echo=FALSE}
# for(i_feature in c(features_TP, features_FP)) {
#   ind_feature <- seq_len(20) == i_feature
#   for(r in tb_sim$R) {
#     fit_original_data <- glmnet::cv.glmnet(x = cbind(tb_sim$x_samples[[r]][, ind_feature],
#                                                      tb_sim$x_samples[[r]][, !ind_feature] %>% 
#                                                        apply(1, MMUPHin:::TSS) %>% 
#                                                        t()) %>% 
#                                              apply(2, function(x) (x - mean(x)) / sd(x)),
#                                            y = tb_sim$y[[r]],
#                                            family = "binomial")
#     load(paste0(dir_output, "samples_for_testing_",
#                 r, "_", i_feature, ".RData"))
#     l_fit_random_data <- (1:100) %>% 
#       future.apply::future_lapply(
#         function(i_dataset)
#           glmnet::cv.glmnet(x = cbind(samples_for_testing[[i_dataset]][, ind_feature],
#                                       samples_for_testing[[i_dataset]][, !ind_feature] %>% 
#                                         apply(1, MMUPHin:::TSS) %>% 
#                                         t()) %>% 
#                               apply(2, function(x) (x - mean(x)) / sd(x)),
#                             y = tb_sim$y[[r]],
#                             family = "binomial")
#       )
#     result_CRT <- list(fit_original_data = fit_original_data,
#                        l_fit_random_data = l_fit_random_data)
#     save(result_CRT, file = paste0(dir_output, "result_CRT_", r, "_",
#                                    i_feature, ".RData"))
#   }
# }
```

```{r summarize, echo=FALSE}
extract_p <- function(result, s = "lambda.min") {
  beta_data <- glmnet:::coef.cv.glmnet(result$fit_original_data, s = s)[2]
  beta_CR <- vapply(result$l_fit_random_data,
                    function(i_result)
                      glmnet:::coef.cv.glmnet(i_result, s = s)[2],
                    0.0)
  list(p = 1 - mean(abs(beta_data) > abs(beta_CR)),
       beta_data = beta_data,
       beta_CR = beta_CR)
}
tb_summary <- tidyr::expand_grid(
  feature = c(features_TP, features_FP),
  R = tb_sim$R) %>%
  dplyr::mutate(i_result = seq_len(dplyr::n())) %>%
  dplyr::left_join(
    tb_sim, by = "R"
  ) %>%
  dplyr::group_by(i_result) %>%
  dplyr::mutate(result = {
    load(paste0(dir_output, "result_CRT_", R, "_",
                feature, ".RData"))
    result_CRT %>% list()
  }) %>%
  dplyr::ungroup()
#
tb_summary <- tb_summary %>%
  dplyr::mutate(
    summary_min =
      result %>% purrr::map(extract_p, s = "lambda.min"),
    summary_1se =
      result %>% purrr::map(extract_p, s = "lambda.1se")
  ) %>%
  dplyr::mutate(
    p_min = summary_min %>%
      purrr::map_dbl("p"),
    p_1se = summary_1se %>%
      purrr::map_dbl("p")
  )

tb_summary %>%
  dplyr::mutate(TP = feature %in% features_TP) %>%
  dplyr::mutate(feature = factor(feature, levels = c(features_TP, features_FP))) %>%
  tidyr::pivot_longer(cols = c(p_min, p_1se),
                      names_to = "criteria",
                      values_to = "p_CRT") %>%
  ggplot(aes(x = feature, y = p_CRT, color = criteria,
             shape = TP)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.25), size = 3)
```

* The CRT scheme does not require features to be normalized, but the fitting model might?

* Zero-inflatedness of features can impact performance, and sampling efficiency.

* Model (both $X$ and $Y$) misspecification?