---
title: "Distilled CRT simulation proof of concept"
author: "Siyuan Ma"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/n/janson_lab/lab/sma/CRT_microbiome/", 
                     echo=FALSE)
library(magrittr)
library(ggplot2)
```

## Distilled CRT
* Goal: test for $Y \perp X_j | \frac{X_{-j}}{\sum X_{-j}}$
* Full CRT
    * Resample $X_j^r$ based on $X_j | \frac{X_{-j}}{\sum X_{-j}}$
    * Fit $Y \sim X_j, \frac{X_{-j}}{\sum X_{-j}}$ and 
      $Y \sim X_j^r, \frac{X_{-j}}{\sum X_{-j}}$. The latter provide null
      distribution.
* Distilled CRT
    * Fit $Y \sim \frac{X_{-j}}{\sum X_{-j}}$ to obtain 
      $\hat{Y} = \hat{E} Y|\frac{X_{-j}}{\sum X_{-j}}$
    * Resample $X_j^r$ based on $X_j | \frac{X_{-j}}{\sum X_{-j}}$
    * Fit $Y - \hat{Y} \sim X_j$ and 
      $Y - \hat{Y} \sim X_j^r$. 
      The latter provide null distribution.

```{r summarize results, echo=FALSE}
dir_output <- "/n/janson_lab/lab/sma/CRT_microbiome/results/dCRT_proof/"
batchtools::loadRegistry(file.dir = "r_batchtools_reg/dCRT_proof/", writeable = FALSE)
load(file = paste0(dir_output, "tb_job_testing.RData"))

extract_p <- function(result) {
  beta_data <- result$fit_original_data$coef[2]
  beta_CR <- vapply(result$l_fit_random_data,
                    function(i_result)
                      i_result$coef[2],
                    0.0)
  list(p = 1 - mean(abs(beta_data) > abs(beta_CR), na.rm = TRUE),
       beta_data = beta_data,
       beta_CR = beta_CR)
}
# 
# tb_result <- tb_job %>%
#   dplyr::filter(i_job %in% batchtools::findDone()$job.id)
# 
# l_results <- list()
# l_summary <- list()
# for(i_row in seq_len(nrow(tb_result))) {
#   if(i_row %% 20 == 0) print(i_row)
#   load(paste0(dir_output, "fit_", tb_result$i_job[i_row], ".RData"))
#   l_results[[i_row]] <- result_CRT
#   l_summary[[i_row]] <- extract_p(result_CRT)
# }

# tb_result <-  tb_result %>%
#   dplyr::mutate(
#     summary = l_summary
#   ) %>%
#   dplyr::mutate(
#     p = summary %>%
#       purrr::map_dbl("p")
#   )
# save(tb_result, file = paste0(dir_output, "tb_result.RData"))
load(paste0(dir_output, "tb_result.RData"))
tb_result_distilled <- tb_result
load("results/CRT_simulation/tb_result.RData")
tb_result <- rbind(
  tb_result %>%
    dplyr::rename(p = p_min,
                     summary = summary_min) %>% 
    dplyr::mutate(method = "full") %>% 
    dplyr::select(-p_1se, -summary_1se),
  tb_result_distilled %>% 
    dplyr::mutate(method = "distilled")
) %>% 
  dplyr::mutate(method = factor(method, levels = c("distilled", "full")))
```


```{r summarize, echo=FALSE}
features_TP <- tibble::tibble(
  mean = (1 - tb_result$params_x[[1]]$pi0) * exp(tb_result$params_x[[1]]$mu)
) %>%
  dplyr::mutate(feature = seq_len(dplyr::n())) %>%
  dplyr::arrange(-mean) %>%
  {.[c(1, 11), ]$feature}
features_FP <- tibble::tibble(
  mean = (1 - tb_result$params_x[[1]]$pi0) * exp(tb_result$params_x[[1]]$mu)
) %>%
  dplyr::mutate(feature = seq_len(dplyr::n())) %>%
  dplyr::arrange(-mean) %>%
  {.[c(2, 12), ]$feature}

tb_result %>%
  dplyr::mutate(feature = factor(feature_test, levels = c(features_TP, features_FP)),
                n = factor(n, levels = c(50, 100, 200))) %>%
  dplyr::filter(penalize_method == "ridge",
                sampling_method == "sequential") %>% 
  ggplot(aes(x = n, y = p, 
             color = method)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.25), size = 3) +
  facet_grid(feature ~ signal_strength) + 
  ggtitle("ridge fits")

tb_result %>%
  dplyr::mutate(feature = factor(feature_test, levels = c(features_TP, features_FP)),
                n = factor(n, levels = c(50, 100, 200))) %>%
  dplyr::filter(penalize_method == "lasso",
                sampling_method == "sequential") %>% 
  ggplot(aes(x = n, y = p, 
             color = method)) +
  geom_point(position = position_jitterdodge(jitter.width = 0.25), size = 3) +
  facet_grid(feature ~ signal_strength) + 
  ggtitle("lasso fits")
```