---
title: "Simple CRT"
author: "Siyuan Ma"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
---
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = ".../")
```

```{r setup2, include=FALSE}
library(magrittr)
library(ggplot2)
smar::sourceDir("R/")
set.seed(1)
```

# Agenda

* SparseDOSSA model, importantly its copula component
* M-H sampling for CRT
* Results diagnostics

# The SparseDOSSA model

Let $(p_1, p_2, \dots, p_j)$ be relative abundances of the $m$ 
microbial features (omitting sample index). SparseDOSSA notes that the joint 
likelihood can be decomposed into 
\begin{align}
f(p_1, p_2, \dots, p_m) &= f(u_1, u_2, \dots, u_m) \times \prod_{j = 1}^m f(p_j)
\end{align}
where $u_j = F(p_j)$ has marginal unifrom distribution. The motivation for this 
approach is to separate the modelling for $p_j$ into univariate marginals, and
a correlation structure captured in $f(u_1, u_2, \dots, u_m)$.

In practice, we assume parametric models for both the marginals of $p_j$ and 
the joint of $u_1, u_2, \dots, u_m$. Specifically,

* $p_j \sim$ zero-inflated logit normal

* $u_1, u_2, \dots, u_m$ are cumulatives of a joint normal 
+ That is, $f(u_1, u_2, \dots, u_m) = f(g_1, g_2, \dots, g_m) /prod_{j = 1}^m f(g_j)$
where $g_j$ form multivariate Gaussian.
+ So that distribution for $u$ is fully parameterized by a correlation matrix 
$\mathbf{\rho}$).

* Not ideal due to zero-inflation? Should sparsity be imposed on 
$\mathbf{\rho}$?

Estimated parameters are plugged into likelihoods for M-H sampling.

# M-H sampling

Noting that 

\begin{align}
f(p_j | p_{-j}/\sum p_{-j}) &= \frac{f(p_1, p_2, \dots, p_m)}{p_{-j}/\sum p_{-j}} \\
&\propto f(p_1, p_2, \dots, p_m) \\
&= f(u_1, u_2, \dots, u_m) \times \prod_{j = 1}^m f(p_j) \\
&= \frac{f(g_1, g_2, \dots, g_m)}{\prod_{j = 1}^m f(g_j)} \times \prod_{j = 1}^m f(p_j)
\end{align}

This can be used as the kernel for M-H proposal. In practice, proposal is 
realized through $g_j$ $\rightarrow$ $g^*_j$ (normal sampling) 
$\rightarrow u^*_j \rightarrow p_j^* \rightarrow p_{-j}^*$.

* Need to run MCMC chain per-sample, per-covariate?

* Again, zero-inflation causes issues? For example, $p \rightarrow u$ when $p$ 
is zero?

# Real data results

Analyzed one of the gut microbiome cohorts in my meta-analysis project. ~100 
samples and features post filtering. First load data and estimate the parameters.

```{r SparseDOSSA fits}
load("data/genera.RData")
physeq <- physeq_genera %>% 
  phyloseq::subset_samples(dataset_name == "MucosalIBD") %>% 
  smar::prune_taxaSamples(flist_taxa = genefilter::kOverA(k = 5, A = 1))
mat_X_count <- smar::otu_table2(physeq)
n_i <- apply(mat_X_count, 2, sum)
lFit_featureParam <- lapply(1:nrow(mat_X_count), function(i_feature) {
  SparseDOSSA2:::estimate_featureParam(
    y_ij = mat_X_count[i_feature, ], 
    n_i = n_i,
    control = list(maxiter.outer = 1000,
                   maxiter.inner = 10000,
                   reltol.outer = 1e-8,
                   factr.inner = 1e7))
})
dfFit_featureParam <- Reduce("rbind", lapply(lFit_featureParam, 
                                             function(x) x$theta)) %>% 
  as.data.frame()
rownames(dfFit_featureParam) <- rownames(mat_X_count)
mat_X <- Reduce("rbind", 
                lapply(
                  lFit_featureParam, 
                  function(x) x$hidden_param$mu_posterior_ij))
mat_X <- exp(mat_X)/(1 + exp(mat_X))
mat_X[mat_X_count == 0] <- 0
dimnames(mat_X) <- dimnames(mat_X_count)
fit_C <- SparseDOSSA2:::estimate_C(feature_abd = mat_X, 
                                   control = list(method = "spearman",
                                                  random = TRUE,
                                                  R = 50))
Rho <- copula::getSigma(fit_C)
dimnames(Rho) <- list(rownames(mat_X), rownames(mat_X))
hist(Rho[lower.tri(Rho)])
```

Run CRT chains on E. coli, known to be associated with disease.

```{r run MCMC on E. coli in one sample}
# First shuffle sample rows to make E. coli come first
name_j <- "k__Bacteria|p__Proteobacteria|c__Gammaproteobacteria|o__Enterobacteriales|f__Enterobacteriaceae|g__|NA|NA"
name_j_minus <- setdiff(rownames(mat_X), name_j)
mat_X <- mat_X[c(name_j, name_j_minus), ]
dfFit_featureParam <- dfFit_featureParam[c(name_j, name_j_minus), ]
Rho <- Rho[c(name_j, name_j_minus), c(name_j, name_j_minus)]
# Run MCMC on the first sample
R <- 2000
K <- 5
# l_ps <- list()
# for(i_sample in 1:ncol(mat_X)) {
#   if(i_sample %% 10 == 0) print(i_sample)
#   l_chain <- metro(p = mat_X[, i_sample],
#                    mu = dfFit_featureParam$mu,
#                    sigma = sqrt(dfFit_featureParam$sigma2),
#                    pi = 1- dfFit_featureParam$pi,
#                    Sigma = Rho, 
#                    K = K, R = R)
#   save(l_chain, file = paste0("results/simple_CRT/",
#                               i_sample,
#                               "_chains.RData"))
#   p_chains <- l_chain %>% 
#     purrr::map2_dfr(
#       1:K,
#       function(chain, k) {
#         data.frame(k = k,
#                    p = chain[-1] %>% 
#                      purrr::map_dbl(~.x[["p"]][1]),
#                    R = 1:R)
#       })
#   l_ps[[i_sample]] <- p_chains
#   
#   traceplot <- p1_chains[sample.int(n = nrow(p1_chains)), ] %>% 
#     ggplot(aes(x = R, y = p, color = as.factor(k))) +
#     geom_point() + geom_line()
#   ggsave(filename = paste0("results/simple_CRT/",
#                            i_sample,
#                            "_chains_trace.pdf"),
#          traceplot, width = 8, height = 4)
# }
# save(l_ps, file = "results/simple_CRT/all_ps.RData")
load("results/simple_CRT/all_ps.RData")
```

Generate some diagnositcs

```{r diagnostics}
# Traceplot
l_ps[[1]][sample.int(n = nrow(l_ps[[1]])), ] %>% 
  ggplot(aes(x = R, y = p, color = as.factor(k))) +
  geom_point() + geom_line()
# Acceptance rates
rates <- c()
for(i_sample in 1:ncol(mat_X)) {
  load(paste0("results/simple_CRT/",
              i_sample,
              "_chains.RData"))
  rate <- l_chain %>% 
    purrr::map(~.x[-1] %>% purrr::map_lgl(~.x$accept)) %>% 
    unlist() %>% 
    mean()
  rates <- c(rates, rate)
}
data.frame(sample = 1:ncol(mat_X), rates = rates) %>% 
  ggplot(aes(x = as.factor(sample), y = rates)) +
  geom_bar(stat = "identity")
# Are they sampling zeros?
l_ps %>% 
  purrr::map_dbl(~mean(.x$p == 0)) %>% 
  data.frame(MCMC_perc_zeros = .) %>% 
  dplyr::mutate(is_zero = mat_X[1, ] == 0) %>% 
  ggplot(aes(x = is_zero, y = MCMC_perc_zeros)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.5))
```

Lastly, perform basic testing on these

```{r testing}
data_test <- smar::sample_data2(physeq)[, "disease", drop = FALSE] %>% 
  cbind(data.frame(t(mat_X))) %>% 
  dplyr::mutate(disease = factor(disease, levels = c("control", "CD")))
fit_realdata <- glm(disease~., data = data_test, family = binomial(), 
                    control = list(maxit = 300))
# pvalue distributions across features
fit_realdata %>% summary %>% coef %>% as.data.frame(check.names = FALSE) %>% 
  ggplot(aes(x = `Pr(>|z|)`)) + geom_histogram()

# pvals_null <- c()
L <- 100
# mat_X_replace <- mat_X
# for(l in 1:L) {
#   mat_X_replace[1, ] <- l_ps %>% 
#     purrr::map_dbl(~.x$p[501:R] %>% 
#                      sample(size = 1))
#   data_null <- smar::sample_data2(physeq)[, "disease", drop = FALSE] %>% 
#     cbind(data.frame(t(mat_X_replace))) %>% 
#     dplyr::mutate(disease = factor(disease, levels = c("control", "CD")))
#   fit_null <- glm(disease~., data = data_null, family = binomial(), 
#                   control = list(maxit = 300))
#   pvals_null <- c(pvals_null, coef(summary(fit_null))[2, 4])
# }
# save(pvals_null, file = "results/simple_CRT/pvals_null.RData")
load("results/simple_CRT/pvals_null.RData")
hist(pvals_null)
```